Recruiting Chapter

### 2020-12-18 manuscript.Rmd
draft changes manuscript

The motivation for these analyses is grounded within the sociology of higher education, particularly two theoretical and empirical syntheses developed by Mitchell Stevens and colleagues [@RN2377; @RN4517]. First, @RN2377 introduces metaphors that describe how sociology has conceived the functions of higher education. The stratification tradition conceives of higher education as a "sieve" -- an instrument that strains and separates -- that allocates people to occupational and social positions. This metaphor builds on "Max Weber's core insight that education has a dual character" [@RN2377, p. 129], on one hand, serving social mobility by redistributing access to privileged positions and, on the other, serving social reproduction by granting privileged families disproportionate access to universities that confer privileged positions. The institutional theory tradition in sociology conceives of higher education as a "hub" that connects, pollinates, and legitimizes "prominent institutional sectors of modern societies" (e.g., professions, the philanthropic sector, corporate sector).  As a hub, higher education is "simultaneously a powerful and fragile social institution. On the one hand, higher education [is powerful because it] connects and reciprocally blesses various forms of privilege. Elite groups (e.g., high-status families, professions, politicians, scientists, intellectuals) and those aspiring to be a part of them use higher education to certify their legitimacy" [@RN2377, p. 137]. On the other hand, higher education is fragile because it depends on "patronage as a primary source of revenue."

@@

# What Do We Know About Off-campus Recruiting?

<!-- The "enrollment funnel," shown in Figure X, describes stages in the student recruitment process (e.g., prospects, inquiries, applicants, accepted applicants, and enrolled students). -->

This section provides context for our analyses by, first, situating off-campus recruiting within the broader set of marketing and recruiting interventions. Second, we review what is known about off-campus recruiting from market research and from academic scholarship, with a particular focus on scholarship that examines the relationship between private high schools and colleges/universities.

Figure \@ref(fig:em-funnel) shows the "enrollment funnel," a conceptual heurisitc used by the enrollment management industry, which describe the stages in the student recruitment process (e.g., prospects, inquiries, applicants, accepted applicants, and enrolled students). For example, “prospects” are “all the potential students you would want to attract to your institution” [@RN4322]. “Inquiries” are prospects that contact the institution, including those who respond to a solicitation and those who reach out on their own. The enrollment management industry uses the enrollment funnel to inform interventions that target specific stages of the funnel.

Table X, reproduced from @RN4664, shows the the percentage of undergraduate recruitment budget allocated to different marketing and recruiting activities. ^[table drawn directly from @RN4664 figure 9; based on a convenience sample of 45 four-year non-profit institutions and 21 four-year public institutions]. Institutions identify undergradaute prospects largely by purchasing "student lists" from College Board, ACT, and other vendors. Institutions utilize advertising (traditional and digital) and social media to solicit inquiries and to create positive "buzz" amongst prospects [@RN4411][CITE OTHER].  Given the the rise in “stealth applicants” who do not inquire before applying (Dupaul & Harris, 2012), advertising and social media enables universities to tell their story to prospects who do not wish to be contacted. Once prospects and inquiries are identified, they are targeted with face-to-face and remote recruiting interventions designed to solit applications and deepen engagement. Face-to-face recruiting interventions include on-campus visits by prospects and off-campus recruiting visits by admissions representatives to high schools, community colleges, college fairs, etc. Remote recruiting interventions include postcards, brochures, email, text messages, and targeted social media.
<!-- QUESTIONS ABOUT THE RNL 2020 COST OF RECRUITING AN UNDERGRADUATE STUDENT REPORT, FIGURE 9

@@

What do we know about off-campus recruiting from previous research? As is true for most aspects of enrollment management, much of what we know about off-campus recruiting comes from consulting firms (e.g., Ruffalo-Noel Levitz, EAB), professional associations (e.g., NACAC), and from practioner-oriented publications. Market research conceives of off-campus recruiting visits as a means of identifying prospects, deepening engagement with prospects already being targeted through mail/email, and maintaining relationships with guidance counselors at “feeder schools” [@RN4323; @RN4402; @RN4664]. With respect to expenditure, @RN4664 found that private (non-profit) 4-yr institutions spent an average of 17% of their undergraduate marketing/recruiting budget on "travel" to high schools and college fairs (as shown in Table X), a higher percentage of budget allocation than any other marketing/recruiting activity. Public institutions spent an average of 16% of their budget on travel, second only to "prospective student communications" at 17%. An emergent trend over the past decade -- partially a response to public universities seeking nonresident students -- has been the growth of "regional recruiters" who target specific metropolitan areas in the US and abroad. These regional recruiters may be college/university employees or they may be independent contractors who live in the metropolitan area they recruit.

A sociological case-study literature develops helpful insights about the mechanisms and functions of off-campus recruiting visits. @RN4324 analyzes visits from the perspective of students at two racially and socioeconomically diverse public high schools. @RN4324 found that high school visits -- including college fairs, instant decision events, and small-group representative visits -- influenced where students applied and where they enrolled. This finding was strongest for first-generation students and under-represented students of color, who often reported that "school counselors had low expectations for them and were too quick to suggest that they attend community college." This trust vacuum created an opportunity for colleges because these students were drawn to colleges that connected with them and "made them feel wanted."  @RN4324 describes a high-achieving, first-generation, African American student who was admitted by a highly selective liberal arts college but chose to attend a less selective college that "seemed to want him more." By contrast, affluent students with college educated parents were less taken by such overtures and more concerned with college prestige.

Of particular importance for this chapter, @RN3519 and @RN4321;@RN4407 highlight the relational function of visits.  @RN3519 provides an ethnography of the admissions office at a selective private liberal arts college, highlights the relational function of visits. During the autumn "travel season," admissions officers visited selected high schools across the country "to spread word of the institution and maintain relationships with guidance counselors" (p. 53-54) because "the College's reputation and the quality of its applicant pool are dependent upon its connections with high schools nationwide" {@RN3519} [p. 54]. The College tended to visit the same schools year after year because recruiting depends on long-term relationships with high schools. The high schools they visited tend to be affluent schools -- in particular, private schools -- that enroll high-achieving students who can afford tuition and had the resources and motivation to host a successful visit. Whereas @RN4402 highlights the effect of recruiting visits on inquiries and enrollees, findings from @RN3519 suggest that the College may have valued recruiting visits primarily as a means of maintaining relationships with guidance counselors. From this perspective, recruiting visits may affect outcomes such as inquiries, applications, and matriculation through their affect on high school guidance counselors. The logic is that a guidance counselor who views a college favorably will steer students to the college.

@RN4321 analyzed recruiting from the perspective of an elite private boarding school in order to understand "how such schools continue to get comparatively under-qualified students into top colleges and universities." The answer to this question begins by thinking about the goals of colleges, which are represented by admissions officers, and the goals of private high schools, which are represented by guidance counselors. Colleges want high achieving students who can pay tuition and donate. Additionally, drawing from @@RN3519, colleges want a class composed of "interesting characters" whose curricular and extracurricular strengths meet the needs of different campus constituents (e.g., academic majors, the athletic department, clubs, etc.). Colleges also want low acceptance and high yield rates to move up the rankings. Private high schools want to send _all_ students to the best college possible. Here, high school guidance counselors face "the pressure of making sure their school seems worth it – that...paying some $40,000+, really does aid students in the college process" [FROM CHAPTER. FIND PG NUMBER]  The challenge counselors face is "some of these students are slightly better than others. These students will likely get into more than one school -- but they can only attend one. And this will lower the chances of your "second-best" students getting into top schools" [@RN4407] [p. 173-174].

"Luckily," @RN4321 [pg. xx] writes, "the problem for elite boarding schools matches up quite nicely with the problem faced by elite colleges." Specifically, college admissions officers receive applications from many outstanding students. But "These outstanding students will also be outstanding to Princeton, Yale, Stanford, and everywhere else. How do you know the ones you pick will attend your school? you can't quite trust applicants, as they are all likely to tell you how much they want to go to your school. And if students you accept go somewhere else, there's not much you can do. But you can get better information -- information you want -- from their high school" [@RN4407] [pg. 173?].

@RN4407 argues that the desire by colleges for trustworthy information about applicant intentions creates an opportunity for high school counselors to advocate on behalf of their students. This opportunity depends on guidance counselors having personal relationships with university admissions offices and on having small enough caseloads to advocate for each student individually. To explain how the process works, @RN4407 describes two hypothetical students -- Susan and Billy -- who both apply to Harvard and Yale.  Susan is a shoo-in at both universities, but wants to attend Harvard. Billy has a weaker academic record than most Ivy League students, but has strong extracurricular activities. The guidance counselor tells Harvard that Susan wants to attend Harvard. Next, he informs Yale that Susan will choose Harvard, but Billy loves Yale and has great "character" and extracurricular activities. In the end, Harvard rejects Billy and accepts Susan (decreasing acceptance rate, increasing yield). Yale rejects Susan and accepts Billy (decreasing acceptance rate, increasing yield).

The horsetrading described by @RN4407 depends on a relationship where the college can trust statements made by the high school counselor and vice-versa. This relationship is the product of repeated interactions over many years. @RN4407 writes that colleges "can reward that [high] school for good information and sanction it for bad information" [p. 173]. A high school that makes false statements about applicant intentions faces consequences. The college "might stop taking these telephone calls and ignore the information provided. They may even start accepting fewer students from the school, thinking it is less than an honest place." [CITE BOOK CHAPTER; PG NUMBER]. Because the college and the high school are mutually dependent, however, both "have an incentive to continue with a strong, honest relationship" [CITE BOOK CHAPTER; PG NUMBER].  Such horsetrading may be less pervasive than now than it was in prior decades, less common at non-elite private schools, and also public university admissions officers likely have less authority to engage in these tacit negotiations.

Nevertheless, maintaining a strong relationships is mutually important for the college and the high school.  Both the college admissions counselor and the high school guidance counselor need to tell one another "their story" and relay that story to constituents. The college admissions counselor explains why the college/university is a "special place" that high school students should want to attend. The guidance counselor explains why the high school is a special place, that even students with lower grades have outstanding extracurricular strengths that will benefit the college. [CITES FOR THIS PARAGRAPH]

Circling back to the focus of this chapter, off-campus recruiting visits are necessary for the maintenance of strong relationships that enable colleges and high schools to negotiate and send trustworthy information to one another. Without face-to-face visits, it is less likely that that a college admissions counselor will "take the call" of a guidance counselor. Simultaneously, a recruiting visit between a college and a high school can be conceived as an indicator. First, the fact that the college took time and effort to make the visit suggests that the college wants to enroll students from the high school. Second, the fact that the high school hosted the visit suggests that the high school likely views the college as a desirable destination for some of its students.  Third, the presence of the recruiting visit indicates that the high school and the college have a relationship and suggests the probability of additional interactions (e.g., phone calls).

@@

Each of these marketing/recruiting interventions -- and dozens of other interventions described in this book -- represent organizational behaviors that could be the subject of empirical research. However, the vast majority of research on enrollment management behaviors focuses on the final stages of the enrollment funnel, specifically which applicants are admitted and the use of financial aid "leveraging" to convert admits to enrollees [E.G.] [@RN3536; @RN3544; @RN3523; @RN4131; @RN2241; @RN3564; @RN1948]

.  By contrast, the enrollment management industry expends substantial resources on marketing/recruiting activities that target earlier stages of the enrollment funnel [@RN4664] [GIVE SOME CONCRETE BUDGET NUMBERS INDICATING BALL PARK SPENDING?]. When we consider descriptions of enrollment management practice from investigative journalism [e.g., ], from enrollment management consulting firms and from this volume [e.g., ], it becomes apparent that -- with the notable exception of a small sociological case-study literature [e.g., ] -- the research community has ignores a great number of enrollment management practices that plausibly affect access to higher education. As a consequence, enrollment management remains an opaque industry to policymakers and the public. We argue that developing thoughtful policies about enrollment management depends on researchers collectively devloping empirical literatures that document enrollment management practices and evaluate the effects of these practices on opportunities for students.

>>>>

The motivation for these analyses is grounded within the sociology of higher education, particularly two theoretical and empirical syntheses developed by Mitchell Stevens and colleagues [@RN2377; @RN4517]. @RN2377 introduces metaphors that describe how sociology has conceived the functions of higher education. The stratification tradition conceives of higher education as a "sieve" -- an instrument that strains and separates -- that allocates people to occupational and social positions. This metaphor builds on "Max Weber's core insight that education has a dual character" [@RN2377, p. 129], on one hand, serving social mobility by redistributing access to privileged positions and, on the other, serving social reproduction by granting privileged families disproportionate access to universities that confer privileged positions. The institutional theory tradition in sociology conceives of higher education as a "hub" that connects, pollinates, and legitimizes "prominent institutional sectors of modern societies" (e.g., professions, the philanthropic sector, corporate sector).  As a hub, higher education is "simultaneously a powerful and fragile social institution. On the one hand, higher education [is powerful because it] connects and reciprocally blesses various forms of privilege. Elite groups (e.g., high-status families, professions, politicians, scientists, intellectuals) and those aspiring to be a part of them use higher education to certify their legitimacy" [@RN2377, p. 137]. On the other hand, higher education is fragile because it depends on "patronage as a primary source of revenue."

@@

# What Do We Know About Off-Campus Recruiting?

<!-- 
This section provides context for our analyses by, first, situating off-campus visits within the broader set of marketing and recruiting interventions. Second, we review what is known about off-campus recruiting from market research and from academic scholarship, emphasizing scholarship that examines the relationship between private high schools and colleges/universities.
-->

This section provides context for our analyses.  First, we situate off-campus recruiting within the broader set of marketing and recruiting interventions in higher education and review what market research says about off-campus recruiting. Second, we review empirical scholarship from sociology that considers off-campus recruiting, emphasizing scholarship that considers the relational aspects of recruiting visits and scholarship that considers private high schools.

## Situating Off-Campus Recruiting Within Enrollment Management

The "enrollment funnel" -- depicted in Figure \@ref(fig:em-funnel) -- is a conceptual heuristic that identifies stages in the student recruitment process (prospects, inquiries, applicants, accepted applicants, and enrolled students). “Prospects” are “all the potential students you would want to attract to your institution” [@RN4322]. “Inquiries” are prospects that contact the institution, including those who respond to a solicitation and those who reach out on their own. The enrollment management industry uses the enrollment funnel to inform marketing and recruiting interventions that target specific stages. Most scholarship on enrollment management focuses on the final stages of the enrollment funnel, specifically which applicants are admitted and the use of financial aid "leveraging" to convert admits to enrollees [e.g., @RN3536; @RN3544; @RN3523; @RN4131; @RN2241; @RN3564; @RN1948]. By contrast, the enrollment management industry expends substantial resources on marketing/recruiting activities that target earlier stages of the enrollment funnel [@RN4664]. [SHOULD WE TRY TO ADD SOME CONCRETE NUMBERS ON $ SPENDING AT PARTICULAR UNIVERSITIES?]

Institutions identify undergraduate "prospects" by purchasing "student lists" -- containing contact, demographic, and academic achievement information -- from College Board, ACT, and other vendors. Universities control which prospects contained within a list by filtering on criteria such as zip code, test scores, and high school GPA. Once prospects and inquiries are identified, they are targeted with remote and face-to-face recruiting interventions designed to solicit applications and deepen engagement. Remote recruiting interventions include postcards, brochures, email, text messages, and targeted social media. Face-to-face recruiting interventions include on-campus visits by prospects and off-campus recruiting visits by admissions representatives to high schools, community colleges, college fairs, etc. Institutions utilize advertising (traditional and digital) and social media to solicit inquiries and to create positive "buzz" amongst prospects [@RN4411][CITE OTHER].  Given the the rise in “stealth applicants” who do not inquire before applying (Dupaul & Harris, 2012), advertising and social media enables universities to tell their story to prospects who do not wish to be contacted.

What do we know about off-campus recruiting from previous research? As is true for most aspects of enrollment management, much of what we know about off-campus recruiting comes from consulting firms (e.g., Ruffalo-Noel Levitz, EAB), professional associations (e.g., NACAC), and from practioner-oriented publications. Market research describes off-campus recruiting visits as a means of identifying prospects and deepening engagement with prospects already being targeted through mail/email [@RN4323; @RN4402; @RN4664].

With respect to expenditure, Table \@ref(tab:ug-recruit-budget), reproduced from @RN4664, shows the the percentage of undergraduate recruitment budget allocated to different marketing and recruiting activities. ^[table drawn directly from @RN4664 Figure 9; based on a convenience sample of 45 four-year non-profit institutions and 21 four-year public institutions]. The average public university spent 16% of its recruiting budget on off-campus recruiting visits, compared to 12% of its budget on purchasing student lists and 17% of its budget on "prospective student communications" (e.g., mail, email, text, social media) targeting prospects and inquiries. An emergent trend over the past decade -- partially a response to public universities seeking nonresident students -- has been the growth of "regional recruiters" who target specific metropolitan areas in the US and abroad [@RN4021]. These regional recruiters may be college/university employees or they may be independent contractors who live in the metropolitan area they recruit.

<!-- QUESTIONS ABOUT THE RNL 2020 COST OF RECRUITING AN UNDERGRADUATE STUDENT REPORT, FIGURE 9

@@

## Empirical Scholarship from Sociology

<!-- QUESTIONS ABOUT THE RNL 2020 COST OF RECRUITING AN UNDERGRADUATE STUDENT REPORT, FIGURE 9

A sociological case-study literature develops helpful insights about the mechanisms and functions of off-campus recruiting visits [CITE].

Empirical academic scholarship on off-campus recruiting is mostly limited to a modest number of sociological case studies, which typically analyze off-campus recruiting as part of a broader analysis of enrollment management [e.g., @RN4324; @RN3519; @RN4538; @RN4520; @RN4407]. @RN4324 analyzes visits from the perspective of students at two racially and socioeconomically diverse public high schools. @RN4324 found that high school visits -- including college fairs, instant decision events, and small-group representative visits -- influenced where students applied and where they enrolled. This finding was strongest for first-generation students and under-represented students of color, who often reported that "school counselors had low expectations for them and were too quick to suggest that they attend community college" (p. XXX) This trust vacuum created an opportunity for colleges because these students were drawn to colleges that connected with them and made them feel wanted.  For example, @RN4324 describes a high-achieving, first-generation, African American student who was admitted by a highly selective liberal arts college but chose to attend a less selective college that "seemed to want him more" (p. XXX). By contrast, affluent students with college educated parents were less taken by such overtures and more concerned with college prestige.

@RN3519 and Khan [-@RN4321;-@RN4407] are centrally important to our study because they analyze relationships between college admissions counselors and high school guidance counselors. @RN3519, an ethnography of the admissions office at a selective private liberal arts college, highlights the relational function of visits. Stevens [-@RN3519, p. 54] states that "the College's reputation and the quality of its applicant pool are dependent upon its connections with high schools nationwide." Therefore, during the autumn "travel season," admissions officers visit selected high schools across the country "to spread word of the institution and maintain relationships with guidance counselors" (p. 53-54). The College tended to visit the same "feeder" schools year after year because recruiting depends on long-term relationships with high schools. The high schools they visited tend to be affluent schools -- in particular, private schools -- that enroll high-achieving students who can afford tuition and had the resources and motivation to host a successful visit. Whereas @RN4402 highlights the effect of recruiting visits on inquiries and enrollees, findings from @RN3519 suggest that the College may have valued recruiting visits primarily as a means of maintaining relationships with guidance counselors. From this perspective, recruiting visits may affect outcomes such as inquiries, applications, and matriculation through their affect on high school guidance counselors. The logic is that a guidance counselor who views a college favorably will steer students to the college.

@RN4321 analyzed recruiting from the perspective of an elite private boarding school in order to understand "how such schools continue to get comparatively under-qualified students into top colleges and universities" (p. 98). The answer begins by considering the goals of colleges, which are represented by admissions officers, and the goals of private high schools, which are represented by guidance counselors. Colleges want high achieving students who can pay tuition and donate. They want low acceptance and high yield rates, which are important for rankings. Colleges also want a class composed of "interesting characters" whose curricular and extracurricular strengths meet the needs of important campus constituents (e.g., academic majors, the athletic department, clubs, etc.). Private high schools want to send _all_ students to the best college possible. Here, high school guidance counselors face "the pressure of making sure their school seems worth it – that \ldots paying some $40,000+, really does aid students in the college process" [@RN4321, p. 105].

The challenge faced by high school counselors at elite private schools is that "some of these students are slightly better than others. These students will likely get into more than one school -- but they can only attend one. And this will lower the chances of your 'second-best' students getting into top schools" [@RN4407, pp. 173-174]. "Luckily," Khan [-@RN4321, p. 105] writes, "the problem for elite boarding schools matches up quite nicely with the problem faced by elite colleges." That is, although college admissions officers receive applications from many outstanding students,

> These outstanding students will also be  outstanding to Princeton, Yale, Stanford, and everywhere else. How do you know the ones you pick will attend your school? You can't quite trust applicants, as they are all likely to tell you how much they want to go to your school. And if students you accept go somewhere else, there's not much you can do. But you can get better information -- information you want -- from their high school. And you can reward that school for good information and sanction it for bad information [@RN4407, p. 173].

 And you can reward that school for good information and sanction it for bad information [@RN4407, p. 173].

This desire by colleges for trustworthy information about applicant intentions creates an opportunity for high school counselors to advocate on behalf of their students. This opportunity depends on guidance counselors having personal relationships with university admissions offices and on having small enough caseloads to advocate for each student individually.

To explain how high school counselors capitalize on this opportunity, @RN4407 describes two hypothetical students -- Susan and Billy -- who both apply to Harvard and Yale. Susan is a shoo-in at both universities, but wants to attend Harvard. Billy has a weaker academic record than most Ivy League students, but has strong extracurricular activities. The guidance counselor tells Harvard that Susan wants to attend Harvard. Next, he informs Yale that Susan will choose Harvard, but Billy loves Yale and has great "character" and extracurricular activities. In the end, Harvard rejects Billy and accepts Susan, thereby, decreasing acceptance rate and increasing yield. Yale rejects Susan and accepts Billy, thereby, decreasing acceptance rate and increasing yield.

The horsetrading described by @RN4407 depends on a relationship where the college can trust statements made by the high school counselor and vice-versa. This relationship is the product of repeated interactions over many years. A high school that makes false statements about applicant intentions faces consequences. The college "might stop taking these telephone calls and ignore the information provided. They may even start accepting fewer students from the school, thinking it is less than an honest" [@RN4321, p. 106]. Because the college and the high school are mutually dependent, however, both "have an incentive to continue with a strong, honest relationship" [@RN4321, p. 106]. However, such horsetrading may be less pervasive than now than it was in prior decades and may be less common at non-elite private schools. Additionally, college admissions officer from public universities likely have less authority to engage in these tacit negotiations because admissions criteria at public universities face greater public scrutiny than admissions criteria at private colleges and universities.

Even in the absence of horsetrading, maintaining strong relationships is mutually important for the college and for the high school. The college admissions counselor and the high school guidance counselor need to tell one another "their story" and relay that story to constituents [@RN3519]. The college admissions counselor explains why the college/university is a "special place" that high school students should want to attend. The guidance counselor explains why the high school is a special place, that even students with lower grades have outstanding extracurricular strengths that will benefit the college.

To summarize, @RN3519 finds that off-campus recruiting visits are important for the maintenance of strong relationships between a college and a high school. Strong relationships enable colleges and schools to negotiate and send trustworthy information to one another and in the absence of strong relationships it is less likely that that a college admissions counselor will "take the call" of a high school guidance counselor [@RN4321; @RN4407]. Building on these findings, we reason that the presence of a recruiting visit between a college and a high school is an indicator that the college and the high school have a relationship. First, the fact that the college made the effort to visit suggests that the college wants to enroll students from the high school. Second, the fact that the high school hosted the visit suggests that the high school likely views the college as a desirable destination for some of its students.  Third, the presence of the recruiting visit suggests the probability of additional interactions (e.g., phone calls).

@@

Each of these marketing/recruiting interventions -- and dozens of other interventions described in this book -- represent organizational behaviors that could be the subject of empirical research.

When we consider descriptions of enrollment management practice from investigative journalism [e.g., ], from enrollment management consulting firms and from this volume [e.g., ], it becomes apparent that -- with the notable exception of a small sociological case-study literature [e.g., ] -- the research community has ignores a great number of enrollment management practices that plausibly affect access to higher education. As a consequence, enrollment management remains an opaque industry to policymakers and the public. We argue that developing thoughtful policies about enrollment management depends on researchers collectively devloping empirical literatures that document enrollment management practices and evaluate the effects of these practices on opportunities for students.
